---
title: "Computing per-mouse rate-based T/C"
date: "Last compiled `r format(Sys.time(), '%a %b %d %Y, %X %Z')`"
output: 
  html_document:
    theme: sandstone
---

```{r setup, message=FALSE}
here::i_am("code/compute_rtc.Rmd",
           uuid = "3a79ff5f-0820-459a-9f6f-95512b7b1349")
library(here)

library(tidyverse)
library(assertthat)
library(conflicted)
filter <- dplyr::filter
set.seed(2)
```

## Load data

```{r}
log10_finite <- function(vals) ifelse(!is.finite(log10(vals)), NA, log10(vals))
dat <- read_tsv(here('data/tumor_growth_data.tsv'),
                show_col_types = FALSE) |>
    mutate(log10_vol = log10_finite(tumor_vol_mm3))

dat2 <- dat
```

## Setup and functions

```{r}
trimmed_slope <- function(df, measure, start_day = 7, end_day = Inf,
                          min_n_measurements = 2) {

    # Helper function for calculating slope
    get_slope_mini <- function(this_dat, var_expr) {
        var_sym <- rlang::ensym(var_expr)
        
        fitdat <- this_dat |>
            filter(!is.na(!!var_sym), !is.na(study_day)) |>
            arrange(study_day)
        
        if (nrow(fitdat) < 2) return(NA)  # not enough to fit a line
        
        # Create formula directly with quosure
        lm1 <- lm(formula(paste0(rlang::as_name(var_sym), " ~ study_day")), data = fitdat)
        return(coef(lm1)[2])
    }
    
    # calculate slopes for each mouse
    df2 <- filter(df, study_day >= start_day, study_day <= end_day) |>
        group_by(study_id, cell_line, strain, trt, mouse_id) |>
        mutate(n_measurements = n()) |>
        nest() |>
        mutate(slope = map(data, ~ get_slope_mini(this_dat = .x, var = {{ measure }}))) |>
        unnest(c(data, slope)) |>
        rename(measure_tumor_growth_slope = slope)
    
    # calculate control (isotype) means
    ctrl_dat <- select(df2, mouse_id, strain, trt, cell_line,
                       measure_tumor_growth_slope, study_id) |>
        distinct() |>
        filter(trt == "isotype") |>
        group_by(study_id, strain, cell_line) |>
        summarize(mean_isotype_measure_tumor_growth_slope =
                      mean(measure_tumor_growth_slope, na.rm = TRUE),
                  .groups = 'drop')

    left_join(df2, ctrl_dat, by = join_by(study_id, cell_line, strain)) |>
        group_by(study_id, cell_line, strain, trt, mouse_id) |>
        mutate(n_measure_notNA = sum(!is.na({{ measure }}))) |>
        mutate(RTC = 10^((measure_tumor_growth_slope -
                              mean_isotype_measure_tumor_growth_slope) * 21)) |>
        mutate(RTC = if_else(n_measurements < min_n_measurements, NA_real_, RTC)) |>
        select(mouse_id, study_id, cell_line, strain, trt,
               n_measurements, n_measure_notNA,
               measure_tumor_growth_slope, RTC) |>
        distinct()
}
```

We originally mapped response after computing a per-mouse RTC by dividing 
each aPD1 mouse's log slope by the mean isotype log slope.

However, this approach posed a problem that was exemplified by strain CC11.
Specifically, in the aPD1 group, all CC11 F1 mice grew tumors but they very
quickly regressed to zero. For many CC11 aPD1 mice there was a single non-zero 
measurement after day 7, but it went down to zero thereafter.
These tumors were not counted in the original RTC metric, because they only 
had a single data point therefore a line could not be fit to the growth curves.
To address this, we took an empirical approach, reasoning that
tumor growth is dramatically shrinking in these mice 
(call them "rapidly shrinking" tumors). 
First we compiled estimates of tumor growth slope in aPD1 and ISO
groups for all mice. We then took the 10th percentile slopes in each group
and replaced these values as an estimate of slope for 
all of the "rapidly shrinking" tumors.

In the function below, we compute RTC as in 
[Hather et al 2014](https://pubmed.ncbi.nlm.nih.gov/25574127/).
but using a per-mouse aPD1 slope and an isotype group mean slope. 
For any tumors that are "rapidly shrinking" tumors as defined 
above, we replaced the unavailable slope estimate with the empirical 10th
percentiles (as above) and computed the RTC in the same manner.

```{r}
mark_first_postgrowth_zero <- function(df) {
    # Mark the first zero tumor volume that occurs after the tumor has shown growth
    
    is_first_postgrowth_zero <- function(tumor_volumes, days) {
        # Sort data by study day
        sorted_data <- tibble(day = days, volume = tumor_volumes) |>
            arrange(day)
        
        # Create a logical vector for zero volumes
        is_zero <- sorted_data$volume == 0
        result <- rep(FALSE, length(is_zero))
        
        # Case 1: No zeros at all - return all FALSE
        if (!any(is_zero)) return(result)
        
        # Case 2: All zeros - return all FALSE (tumor never grew)
        if (all(is_zero)) return(result)
        
        # Find first positive volume (evidence of tumor growth)
        first_growth_idx <- min(which(sorted_data$volume > 0))
        
        # Check if any zeros occur after first growth
        post_growth_zeros <- which(is_zero & seq_along(is_zero) > first_growth_idx)
        
        # If no zeros after growth, return all FALSE
        if (length(post_growth_zeros) == 0) return(result)
        
        # Mark the first post-growth zero
        result[post_growth_zeros[1]] <- TRUE
        return(result)
    }
    
    # Process all mice data
    df |>
        group_by(study_id, cell_line, strain, trt, mouse_id) |>
        mutate(first_postgrowth_zero = is_first_postgrowth_zero(
            tumor_vol_mm3, study_day
        )) |>
        ungroup() # Explicitly ungroup to avoid unexpected behavior
}
```

```{r}
rtc_replace_rapid_shrinkers <- function(df, slope_var, quantile_df) {
    # Replace slope values for "rapid shrinking" tumors with predetermined
    # quantile values
    #
    # Args:
    #   df: Dataframe of tumor response data with an is_rapid_shrinker column
    #   slope_var: The variable name (unquoted) containing slope values
    #   quantile_df: A dataframe containing replacement values
    
    # Join with the quantile dataframe containing replacement values
    df_with_quantiles <- df |>
        left_join(quantile_df, by = join_by(trt, cell_line))
    
    # Calculate isotype control means using either original slopes
    # or replacement values for rapid shrinkers
    isotype_means <- df_with_quantiles |>
        filter(trt == 'isotype') |>
        ungroup() |>
        mutate(adjusted_slope = if_else(is_rapid_shrinker, q10, {{ slope_var }})) |>
        group_by(strain, cell_line, study_id) |>
        summarize(
            iso_mean_slope = mean(adjusted_slope, na.rm = TRUE), 
            .groups = 'drop'
        )
    
    # Calculate final RTC values for all samples using the study-specific control values
    df_with_quantiles |>
        ungroup() |>
        mutate(adjusted_slope = if_else(is_rapid_shrinker, q10, {{ slope_var }})) |>
        select(study_id, strain, cell_line, trt, mouse_id, adjusted_slope) |>
        left_join(isotype_means, by = join_by(strain, cell_line, study_id)) |>
        mutate(RTC = 10^((adjusted_slope - iso_mean_slope) * 21))
}
```

## Do the RTC analysis

```{r}
# mark the mouse_id for tumors that are "rapidly shrinking" as defined above
has_postgrowth_zero <- mark_first_postgrowth_zero(dat2) |>
    group_by(study_id, strain, trt, cell_line, mouse_id) |>
    summarize(has_pgz = any(first_postgrowth_zero), .groups = 'drop')

# calculate RTC
slopes <- trimmed_slope(dat2, measure = log10_vol)
slopes2 <- ungroup(slopes) |>
    mutate(one_point = n_measure_notNA == 1) |>
    inner_join(has_postgrowth_zero) |>
    mutate(is_rapid_shrinker = one_point & has_pgz)

# get the 10th percentile for computed slopes for each trt group/cell line
qq <- group_by(slopes, trt, cell_line) |>
    summarize(q10 = quantile(measure_tumor_growth_slope, 0.1, na.rm = TRUE),
              .groups = 'keep')

rtc_per_mouse <- rtc_replace_rapid_shrinkers(slopes2, 
                                             measure_tumor_growth_slope,
                                             quantile_df = qq) |>
    arrange(cell_line, strain, trt, mouse_id) |>
    mutate(across(where(is.numeric), ~signif(., digits = 6)))
write_tsv(rtc_per_mouse, file = here('data/rtc_permouse.tsv'))

rtc <- rtc_per_mouse |>
    filter(trt != 'isotype') |>
    group_by(strain, cell_line) |>
    summarize(mean_log_RTC = signif(mean(log(RTC), na.rm = TRUE), digits = 6),
              .groups = 'keep') |>
    arrange(cell_line, strain)
write_tsv(rtc, file = here('data/rtc_avg.tsv'))
```
