---
title: "Mapping in CC12xCC80 CCF1"
date: "Last updated `r format(Sys.time(), '%a %b %d %Y, %X %Z')`"
output:
  html_document:
    code_download: true
---

```{r setup, message=FALSE}
here::i_am("code/map_CC12_CC80_F1N1.Rmd",
           uuid = "d11426c5-dd0b-4416-8208-3704f3a0c11c")
library(here)

library(tidyverse)
library(assertthat)
library(qtl2)
set.seed(1)
```

# Load data

Load objects/data for QTL mapping:
```{r}
pr <- readRDS(here('figshare_data/CC12_CC80_F1N1_probs.rds'))
K <- readRDS(here('figshare_data/CC12_CC80_F1N1_K.rds'))
obj <- readRDS(here('figshare_data/CC12_CC80_F1N1_obj.rds'))
assert_that(check_cross2(obj))
```

Next load the phenotypes
```{r load_pheno}
pheno <- read_tsv(here("data/CC12_CC80_F1N1_pheno.tsv"),
                  show_col_types = FALSE)
```

## Map the phenotypes

QTL map the phenotype.

 * CR = no growth *ever*
 * CR* = complete responder that showed initial growth but regressed to zero
 * PR = partial responder that regressed to <1/2 maximum tumor volume
 * NR = non-responder

```{r mapping1}
pdat <- filter(pheno, trt != 'isotype') |>
  mutate(complete_responder=abbrev %in% c("CR", "CR*")) |>
  as.data.frame() |> 
  column_to_rownames('mouse_id')
pheno_dat <- pdat[, c('log10_tumor_growth_slope', 'complete_responder'), 
                  drop=FALSE]
pheno_dat$complete_responder <- as.numeric(pheno_dat$complete_responder)
```

```{r mapping2, fig.width = 8, fig.height = 4}
out <- scan1(pr, pheno_dat, K)
attr(out, 'sample_size')
plot(out, obj$gmap, lodcolumn = 'complete_responder')
```

```{r, fig.width = 8, fig.height = 4, eval = FALSE}
operm <- scan1perm(genoprobs = pr, pheno = pheno_dat, kinship = K, 
                   n_perm = 1000, cores = 6)
thresh <- summary(operm)['0.05', 'complete_responder', drop = FALSE]
plot(out, obj$gmap, lodcolumn = 'complete_responder', ylim = c(0, 7))
abline(h = thresh, col = 'red', lty = 2)
plot(out, obj$gmap, lodcolumn = 'complete_responder', add = TRUE)
```
