---
title: "Analysis of scRNA-Seq data"
date: "Last updated `r format(Sys.time(), '%a %b %d %Y, %X %Z')`"
output: html_document
---

```{r setup, include = TRUE, message = FALSE}
here::i_am("code/single_cell_process.Rmd", 
           uuid = "65af7b05-1c7a-4c33-83d8-7090c303f451")
library(here)
library(tidyverse)
library(Seurat)
library(harmony)
library(assertthat)
#library(presto)   # for more efficient FindMarkers() wilcox.test
library(Matrix)
library(ggpubr)
```

# Load data

We use the processed data available from NCBI GEO
[link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE274421).

```{r}
mdat <- read_tsv(here("data/GSE274421_mf_scRNAseq_metadata.tsv.gz"),
                 show_col_types = FALSE) |>
    as.data.frame()
x <- as(readMM(here("data/GSE274421_mf_scRNAseq.mtx.gz")), "CsparseMatrix")
rownames(x) <- scan(here('data/GSE274421_mf_scRNAseq_genes.tsv.gz'), what = '')
colnames(x) <- scan(here('data/GSE274421_mf_scRNAseq_barcodes.tsv.gz'), what = '')
rownames(mdat) <- colnames(x)
obj <- CreateSeuratObject(counts = x, meta.data = mdat)
```

# Processing and dimensionality reduction

```{r}
# First, load raw data into Seurat objects and merge into one
# object -- `obj`
responder <- c('B6' = 'intermediate', 'CC075' = 'R', 'CC001' = 'R', 
               'CC002' = 'R', 'CC007' = 'intermediate', 
               'CC080' = 'NR', 'CC079' = 'NR', 'CC036' = 'NR')
obj$responder <- setNames(responder[as.character(obj$strain)], Cells(obj))

cc.genes <- readLines(here("data/regev_lab_cell_cycle_genes_mm.fixed.txt"))
s.genes <- cc.genes[1:43]
g2m.genes <- cc.genes[44:98]

obj <- NormalizeData(obj, normalization.method = "LogNormalize") |> 
    PercentageFeatureSet(pattern = '^mt-', col.name = 'pct_counts_mt') |>
    PercentageFeatureSet(pattern = "^Rpl|^Rps", col.name = 'pct_counts_ribo') |>
    subset(subset = pct_counts_mt < 15 & pct_counts_ribo < 45) |>
    FindVariableFeatures(selection.method = "vst", nfeatures = 2000) |>
    CellCycleScoring(s.features = s.genes, g2m.features = g2m.genes, 
                     set.ident = FALSE) |>
    ScaleData(vars.to.regress = c("nCount_RNA", "pct_counts_mt", 
                                  'pct_counts_ribo', "S.Score", "G2M.Score")) |>
    RunPCA(verbose = FALSE, npcs = 100)
ElbowPlot(obj)
# visual examination of ElbowPlot
num_pc <- 37
ElbowPlot(obj, ndims = 60) + geom_vline(xintercept = num_pc)

obj <- FindNeighbors(obj, reduction = 'pca', dims = 1:num_pc, verbose = FALSE) |>
  FindClusters(verbose = FALSE, resolution = 0.05) |>
  RunUMAP(reduction = 'pca', dims = 1:num_pc, verbose = FALSE)
obj$global_cluster_num_not_batch_corrected <- as.character(obj$seurat_clusters)
```

Use `harmony` to correct across processing date
```{r harmony}
obj <- RunHarmony(obj, group.by.vars = 'dataset', 
    theta = 1, dims.use = 1:num_pc, max_iter = 100) |>
    FindNeighbors(reduction = 'harmony', dims = 1:num_pc) |>
    FindClusters(verbose = FALSE, resolution = 0.1) |>
    RunUMAP(dims = 1:num_pc, reduction = 'harmony')
obj$global_cluster_date_corrected <- as.character(obj$RNA_snn_res.0.1)
```

Clusters given in NCBI GEO data include subcluster numbers as well.
Let's label major clusters with names:
 
```{r naming2}
nm2 <- c('0' = 'monomac', '1' = 'T/NK cells', '2' = 'cDC2', 
         '3' = 'monomac', '4' = 'proliferating monomac', '5' = 'T/NK cells',
         '6' = 'MC38', '7' = 'cDC1', '8' = 'pDC/B cells', 
         '9' = 'CAF', '10' = 'gMDSC')
obj$cluster_name <- setNames(nm2[obj$global_cluster_date_corrected],
                             Cells(obj))
UMAPPlot(obj, label = T, label.size = 5, group.by = 'cluster_name') + 
    NoLegend() + xlab("UMAP 1") + ylab("UMAP 2")
```

Subclustering was conducted in a similar manner.

# Examine subcluster differences

```{r}
theme_this <- function(base_size = 18) { 
    theme_bw(base_size = base_size, base_family = 'Arial') +
    theme(axis.text.x = element_blank(), 
          axis.ticks.x = element_blank(),
          panel.grid = element_blank(),
          panel.spacing = unit(0, 'null'),
          plot.title = ggtext::element_markdown(hjust = 1, size = base_size),
          strip.background = element_rect(fill = "white"),
          strip.text = ggtext::element_markdown(size = base_size))
}

symbols <- list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, Inf), 
                symbols = c("****", "***", "**", "*", "n.s."))

```

Compute signature of IFNg stimulation
```{r}
ifng_genes <- c("Ccl5", "Cd27", "Cd274", "Cd276", "Cd8a", "Cmklr1", "Cxcl9",
                "Cxcr6", "H2-Aa", "H2-Eb1", "H2-T23", "Ido1", "Lag3", "Nkg7",
                "Pdcd1lg2", "Psmb10", "Stat1", "Tigit")
obj <- AddModuleScore(obj, features = list(IFNg = ifng_genes), 
                      name = "IFNg", seed = 1)
noI <- subset(obj, responder != 'intermediate')

```

```{r}
sfrac <- select(obj@meta.data, dataset, treatment, mouseID, strain, 
                responder, cluster) %>%
    group_by(across(-cluster)) %>%
    mutate(N_tot = n()) %>%
    group_by(pick(everything())) %>%
    summarize(n_clust = n(), .groups = 'keep') 
toplot <- filter(sfrac, cluster %in% c(3, 13)) %>%
    mutate(frac = n_clust / N_tot) %>%
    filter(responder != 'intermediate') %>%
    mutate(R = ifelse(responder == 'R', 'Responder', 'Non-responder')) %>%
    mutate(condr = interaction(treatment, responder)) %>%
    mutate(condrF = factor(condr, levels = c('isotype.NR', 'isotype.R',
                                             '\u03b1PD1.NR', '\u03b1PD1.R'))) %>%
    mutate(clusterN = case_when(cluster == 3 ~ "Exhausted CTLs",
                                cluster == 13 ~ "IFN\u03b3<sup>+</sup> CTLs",
                                TRUE ~ 'missing'))

```

```{r}
filter(toplot, cluster == 3) %>%
    ggplot(aes(x = R, y = 100*frac)) +
    geom_violin(aes(fill = R), alpha = 0.4, width = 0.75) +
    scale_discrete_manual(values = c('Responder' = '#4dac26', 
                                     'Non-responder' = '#d01c8b'),
                          aesthetics = c('color', 'fill')) +
    geom_point(aes(color = R), 
               position = position_jitter(width = 0.05, seed = 1)) +
    theme_this(base_size = 24) +
    facet_wrap(~ treatment, strip.position = 'bottom') +
    guides(color = 'none', fill = guide_legend(reverse = TRUE)) +
    stat_compare_means(comparisons = list(1:2, 3:4),
                       symnum.args = symbols,
                       vjust = 0, size = 6) +
    scale_y_log10(expand = expansion(mult = c(0.25, 0.25))) +
    xlab("") + ylab("Fraction cells") +
    ggtitle("Exhausted CTLs (Subcluster 3)")

filter(toplot, cluster == 13) %>%
    ggplot(aes(x = R, y = 100*frac)) +
    geom_violin(aes(fill = R), alpha = 0.4, width = 0.75) +
    scale_discrete_manual(values = c('Responder' = '#4dac26', 
                                     'Non-responder' = '#d01c8b'),
                          aesthetics = c('color', 'fill')) +
    geom_point(aes(color = R), 
               position = position_jitter(width = 0.05, seed = 1)) +
    theme_this(base_size = 24) +
    facet_wrap(~ treatment, strip.position = 'bottom') +
    guides(color = 'none', fill = guide_legend(reverse = TRUE)) +
    stat_compare_means(comparisons = list(1:2, 3:4),
                       symnum.args = symbols,
                       vjust = 0, size = 6) +
    scale_y_log10(expand = expansion(mult = c(0.25, 0.2))) +
    xlab("") + ylab("Fraction cells") +
    ggtitle("IFN\u03b3<sup>+</sup> CTLs (Subcluster 13)")
```

```{r}
q75 <- quantile(noI$IFNg1, 0.75)
noI$IFNg_top25 <- ifelse(noI$IFNg1 > q75, 'hi', 'lo')
noI$mac2526_hiIFNg <- ifelse(noI$IFNg_top25 == 'hi' & 
                                 noI$cluster %in% c(2, 5, 26), 
                             'yes', 'no')
sfrac <- select(noI@meta.data, dataset, treatment, mouseID, strain, 
                responder, mac2526_hiIFNg) %>%
    group_by(across(-mac2526_hiIFNg)) %>%
    summarize(frac_2526_hiIFNg = mean(mac2526_hiIFNg == 'yes'),
              .groups = 'keep')

sfrac %>%
    mutate(R = case_when(responder == 'R' ~ 'Responder',
                         responder == 'NR' ~ 'Non-responder',
                         TRUE ~ 'missing')) %>%
    ggplot(aes(x = R, y = 100*frac_2526_hiIFNg)) +
    geom_violin(aes(fill = R), alpha = 0.4, width = 0.75) +
    scale_discrete_manual(values = c('Responder' = '#4dac26', 
                                     'Non-responder' = '#d01c8b'),
                          aesthetics = c('color', 'fill')) +
    geom_point(aes(color = R), 
               position = position_jitter(width = 0.05, seed = 1)) +
    theme_this(base_size = 24) +
    facet_wrap(~ treatment, strip.position = 'bottom') +
    guides(color = 'none', fill = guide_legend(reverse = TRUE)) +
    stat_compare_means(comparisons = list(1:2, 3:4), size = 6,
                       vjust = 0, symnum.args = symbols,
                       label.x.npc = 'center', label.y.npc = 'top') +
    scale_y_log10(expand = expansion(mult = c(0.15, 0.15))) +
    ylab("Fraction of all cells") +
    coord_cartesian(ylim = c(2.5, 75), clip = 'off')
```


